---
title: "Introduction to TargetSureR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to TargetSureR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  # This vignette documents a workflow whose inputs are multi-gigabyte external
  # downloads and network services, so nothing runs at build time: eval = FALSE
  # here, and purl = FALSE on each chunk so that R CMD check does not execute
  # the tangled script either.
  eval     = FALSE,
  fig.align = "center"
)
```

# Introduction

`TargetSureR` is a modular R toolkit for assessing the off-target risk of
short interfering RNAs (siRNAs). The package combines transcriptome
scanning, curated gene-list enrichment, multi-tissue expression context,
and cross-species ortholog conservation into a single composite-score
framework. A built-in cohort of 94 published human siRNAs ships inside
the package and is used as a normalization reference, so that an
investigator can place a single new candidate on a meaningful absolute
scale rather than relying on within-batch ranking alone.

This vignette walks through a complete first-time use, from installation
through the construction of a local sequence database, evaluation of one
candidate siRNA, and interpretation of the resulting tables and figures.

# What you need before you start

TargetSureR ships no genome, no transcriptome and no expression data. The
package is the analysis logic; the reference data are external, and you
assemble them once. **Read this section before running anything else** — it
is the difference between a working installation and one that silently
produces meaningless numbers.

## The four external resources

| Resource | Size | Needed for | Ships with package? |
|---|---|---|---|
| `BSgenome.Hsapiens.UCSC.hg38` | ~800 MB installed | Mode A scanning | No — Bioconductor |
| `EnsDb` SQLite (Ensembl 113) | ~510 MB | Mode A and Mode B | No — AnnotationHub |
| Tissue expression matrix | ~20 MB prepared | `tissue_safety()` | Simulated example only |
| Ortholog table | ~200 KB | `species_match()` | Simulated example only |

The last two deserve emphasis. `gtex_example` and `ortholog_table_example`
contain **20 genes of invented numbers each**. They exist so that examples
run offline. Scoring a real off-target set against them does not fail with an
error — it returns a table in which almost every gene is unquantified, and
the resulting ranking is arbitrary. Both must be replaced for any analysis
whose output you intend to act on.

Three curated gene lists (`ae_genes`, 35 symbols; `cancer_genes_v2`, 979;
`immune_genes_immport`, 1,794; 2,632 unique in total) *are* bundled in full,
so no download is needed for the `critical`-gene flags. Supply your own via
the `gene_lists` argument of `enrich_annotations()` if your programme uses
different panels.

## A suggested layout

Nothing in the package requires a particular directory. The examples in this
vignette assume one folder holding the prepared files, with an R variable
pointing at it, so that no path is written twice:

```{r data-dir, purl = FALSE}
# Adjust to wherever you keep reference data.
TSR_DATA <- "~/TargetSureR_data"
dir.create(TSR_DATA, recursive = TRUE, showWarnings = FALSE)
```

```
~/TargetSureR_data/
├── EnsDb.Hsapiens.v113.sqlite        # step 2, ~510 MB
├── hg38_3UTR_annodb.rds              # step 3, built locally
├── hg38_5UTR_annodb.rds              # step 3
├── hg38_CDS_annodb.rds               # step 3
├── gtex_median_tpm.tsv               # step 4, prepared from GTEx
└── ortholog_biomart.csv              # step 5, from Ensembl BioMart
```

Everything below writes into `TSR_DATA`. Preparation is a one-time cost of
roughly an hour, most of it download time; afterwards a session loads the
cached files in seconds and needs no network.

## What works without each resource

You do not need all four to get started. The pipeline degrades in known ways:

| You have | You can run | You cannot run |
|---|---|---|
| Package only | `normalize_guide()`, `compute_risk_score()`, column specs | anything requiring sequence scanning |
| \+ BSgenome + EnsDb + annodb | `annotate_sites()`, `enrich_annotations()`, `rank_sirna()`, cohort percentiles | meaningful `tissue_safety()` or `species_match()` |
| \+ expression matrix | `tissue_safety()`, `plot_tissue_heatmap()`, `plot_gene_attention()` | meaningful `species_match()` |
| \+ ortholog table | `species_match()`, all cross-species figures | — |

Note that `rank_sirna()` and its cohort percentiles work as soon as scanning
does: the 94-siRNA `reference_set` is bundled in full and needs no download.

# Installation

## Core dependencies

The package relies on a small set of CRAN packages for tabular and
plotting work:

```{r install-cran, purl = FALSE}
pkgs <- c(
  "dplyr", "tibble", "tidyr", "readr", "rlang",
  "ggplot2", "scales", "ggrepel", "RColorBrewer", "httr2"
)

install.packages(setdiff(pkgs, installed.packages()[, "Package"]))
invisible(lapply(pkgs, library, character.only = TRUE))

```

Bioconductor packages are required for sequence handling and Ensembl
annotation. Because access to bioconductor.org may be unstable in some
networks, a local mirror (Westlake or USTC) is recommended:

```{r install-bioc, purl = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

bioc_pkgs <- c(
  "SeedMatchR", "Biostrings", "ensembldb",
  "AnnotationHub", "AnnotationFilter",
  "GenomicRanges", "GenomicFeatures", "GenomeInfoDb",
  "BiocGenerics", "IRanges", "S4Vectors",
  "biomaRt"            # step 5: ortholog download
)

# Install missing Bioconductor packages without updating existing ones
BiocManager::install(setdiff(bioc_pkgs, installed.packages()[, "Package"]), update = FALSE)
# Load all Bioconductor packages
invisible(lapply(bioc_pkgs, library, character.only = TRUE))

```

`SeedMatchR` supplies the seed-scanning engine used in Mode A, and it needs a
word of its own. CRAN carries version 1.1.1; version 2.0.0, used for the
published analyses, is distributed through the authors' GitHub repository and
requires R >= 4.3.0. If `BiocManager::install("SeedMatchR")` gives you 1.1.1,
or nothing at all on a recent R, install from source:

```{r install-seedmatchr, purl = FALSE}
packageVersion("SeedMatchR")   # check what you actually have

# remotes::install_github("tacazares/SeedMatchR")
```

Both versions expose the functions TargetSureR calls, so 1.1.1 works. Record
which one you used: the seed-matching engine is part of what determines a site
count.

## Step 1: the reference genome

The GRCh38 genome sequence arrives as a Bioconductor annotation package,
`BSgenome.Hsapiens.UCSC.hg38` (about 800 MB installed). The analyses
published with this package used version 1.4.5.

```{r install-bsgenome, purl = FALSE}
# Option A: install through Bioconductor (preferred)
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
packageVersion("BSgenome.Hsapiens.UCSC.hg38")   # 1.4.5 in the published analyses
```

A single-session install of this size fails on some networks. The tarball can
instead be fetched with a browser or download manager, then installed from
wherever it landed:

```{r install-bsgenome-local, eval = FALSE, purl = FALSE}
# Download from the annotation-package index for your Bioconductor release:
#   https://bioconductor.org/packages/release/data/annotation/
# then install the local file:
install.packages(
  file.path(TSR_DATA, "BSgenome.Hsapiens.UCSC.hg38_1.4.5.tar.gz"),
  repos = NULL, type = "source")
```

## Installing TargetSureR

```{r install-pkg, purl = FALSE}
# install.packages("remotes")
remotes::install_github("nishuai/TargetSureR")
library(TargetSureR)
packageVersion("TargetSureR")   # 1.1.0
```

# Preparing the reference data

## Step 2: the Ensembl annotation snapshot

Transcript coordinates come from an `EnsDb` SQLite file, about 510 MB. The
published analyses used **Ensembl release 113**, distributed through
AnnotationHub as record **`AH119325`** ("Ensembl 113 EnsDb for Homo sapiens",
added 2024-10-28).

Pinning a release matters for reproducibility: site counts are relative to a
particular transcript set, so a different release gives different numbers. A
newer release is a legitimate choice, but record which one you used.

```{r ensdb-hub, purl = FALSE}
library(AnnotationHub)
ah <- AnnotationHub()

# Find the record for the release you want
query(ah, c("EnsDb", "Homo sapiens", "113"))

# Retrieve it (first call downloads; later calls read the cache)
edb <- ah[["AH119325"]]
ensembldb::ensemblVersion(edb)   # 113
```

### If the AnnotationHub download fails

On constrained networks this download frequently stops partway and leaves a
truncated file, which then fails to open as a database. The file is a plain
SQLite database served over HTTPS, so it can be fetched directly with a
browser or a resumable download tool and opened by path. This is the route
used to produce the published analyses.

```{r ensdb-manual, purl = FALSE}
# 1. Resolve the download URL for the record (run once, with network access):
#      ah <- AnnotationHub::AnnotationHub()
#      AnnotationHub::hubUrl(ah)          # the hub endpoint
#      mcols(ah["AH119325"])              # record metadata
#
#    Or browse the AnnotationHub record listing and copy the URL for
#    AH119325 from there.
#
# 2. Download it with a resumable client, e.g.
#      curl -L -C - -o EnsDb.Hsapiens.v113.sqlite <url>
#
# 3. Place it in TSR_DATA and open it by path:

ensdb_path <- file.path(TSR_DATA, "EnsDb.Hsapiens.v113.sqlite")
edb <- ensembldb::EnsDb(ensdb_path)

# Confirm you have the release you think you have
ensembldb::ensemblVersion(edb)   # 113
ensembldb::organism(edb)         # "Homo sapiens"
```

Verify the file before relying on it: a truncated download often opens
without complaint and then fails mid-analysis.

```{r ensdb-verify, purl = FALSE}
file.size(ensdb_path) / 1024^2        # expect roughly 510 MB
length(ensembldb::transcripts(edb, columns = "tx_id")$tx_id)  # ~250,000
```

Note that `AnnotationHub` caches into a platform-specific directory
(`tools::R_user_dir("AnnotationHub", "cache")` reports it), which differs
between Windows, macOS and Linux. Keeping your own copy under `TSR_DATA` and
passing `ensdb_path` explicitly avoids depending on that location, and makes
a script portable between machines.

## Step 3: the transcript sequence index (`annodb`)

Sequence-mode scanning (Mode A) needs transcript *sequences*, not just
coordinates. `build_annodb()` combines the two inputs above into the object
`annotate_sites()` expects, one per feature class: 3'UTR, 5'UTR or CDS.

Building takes roughly 25 seconds per class, so cache the result and reload
it in later sessions.

```{r build-annodb, purl = FALSE}
annodb_3utr <- build_annodb("3UTR", ensdb_path = ensdb_path)
annodb_5utr <- build_annodb("5UTR", ensdb_path = ensdb_path)
annodb_cds  <- build_annodb("CDS",  ensdb_path = ensdb_path)

saveRDS(annodb_3utr, file.path(TSR_DATA, "hg38_3UTR_annodb.rds"), compress = "xz")
saveRDS(annodb_5utr, file.path(TSR_DATA, "hg38_5UTR_annodb.rds"), compress = "xz")
saveRDS(annodb_cds,  file.path(TSR_DATA, "hg38_CDS_annodb.rds"),  compress = "xz")
```

With Ensembl 113 this yields 19,022 3'UTR, 18,973 5'UTR and 19,355 coding
sequences. Four filters produce those counts, and because every downstream
count is relative to this transcript set, all four have to match for a
published number to be reproducible:

1. protein-coding transcripts only;
2. sequence names restricted to `1:22`, `X`, `Y`, `MT`, dropping scaffolds
   and patch contigs;
3. one transcript per gene, the longest **within that feature class**;
4. sequences shorter than 8 nt dropped, a 7-nt seed being unlocatable in them.

Filter 3 has a consequence worth noting: a gene can contribute a different
representative transcript to each of the three classes, so a per-gene site
count summed across classes is an upper bound on the number of distinct
positions. `?build_annodb` documents all four, and each is adjustable.

Three feature classes are built because `risk_score` derives jointly from
`region` (CDS = 5, 3' UTR = 3, 5' UTR = 1) and `match_type` (full
complementarity = +2). Scanning only one class confines hits to a single
region, and the four risk-score classes (1, 3, 5, 7) cannot all be observed.
The built-in `reference_set` was built by scanning all three for each of the
94 reference siRNAs.

After the initial build, later sessions only load:

```{r load-annodb, purl = FALSE}
annodb_3utr <- readRDS(file.path(TSR_DATA, "hg38_3UTR_annodb.rds"))
annodb_5utr <- readRDS(file.path(TSR_DATA, "hg38_5UTR_annodb.rds"))
annodb_cds  <- readRDS(file.path(TSR_DATA, "hg38_CDS_annodb.rds"))
length(annodb_3utr$seqs)   # 19022
```

## Step 4: tissue expression

Steps 4 and 5 prepare annotation tables that are joined to results *after*
scanning. The downloads are independent of your siRNAs, so they can be done
now; the code that consumes them appears under
\dQuote{Downstream analysis} below. Step 5 additionally needs the gene list
your own scan produces, so its download is best run once you have results.

`tissue_safety()` and the tissue figures need a median-expression matrix:
first column `gene_symbol`, one numeric column per tissue. Without one, the
bundled `gtex_example` is used — 20 genes of simulated values, enough for the
examples below and useless for analysis.

The published analyses used the GTEx v10 gene median TPM file
(`GTEx_Analysis_v10_RNASeQCv2.4.2_gene_median_tpm.gct`, 59,033 genes across
68 tissues), downloaded from the GTEx portal's datasets page
(<https://gtexportal.org/home/downloads/adult-gtex>) under **Bulk tissue
expression**. GTEx requires no registration for these summary files.

The GCT format needs a small conversion: three header lines, then one row per
versioned Ensembl gene ID with the symbol in column 2. Several IDs can map to
one symbol, so rows are collapsed per symbol. The published analyses took the
**per-tissue maximum**, which keeps the highest reported value for a symbol
rather than diluting it with unexpressed paralogous entries.

```{r gtex-prepare, purl = FALSE}
gct <- file.path(TSR_DATA, "GTEx_Analysis_v10_RNASeQCv2.4.2_gene_median_tpm.gct")

# Header: line 1 version tag, line 2 dimensions, line 3 column names
tissues <- strsplit(readLines(gct, n = 3)[3], "\t")[[1]][-(1:2)]
raw <- readr::read_tsv(gct, skip = 3, col_names = c("gene_id", "gene_symbol", tissues),
                       show_col_types = FALSE)

gtex <- raw |>
  dplyr::filter(!is.na(.data$gene_symbol), nzchar(.data$gene_symbol)) |>
  dplyr::select(-"gene_id") |>
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ max(.x, na.rm = TRUE)),
                   .by = "gene_symbol")

readr::write_tsv(gtex, file.path(TSR_DATA, "gtex_median_tpm.tsv"))
dim(gtex)   # 57853 x 69  (gene_symbol + 68 tissues)
```

Record the source file's checksum alongside the output, so a result can be
traced to the exact input:

```{r gtex-checksum, purl = FALSE}
tools::md5sum(gct)
```

In later sessions, load the prepared file and pass it wherever the pipeline
takes an expression matrix. `sites` is produced by the scanning step covered
below, so this is the call you will make once you reach it:

```{r gtex-use, purl = FALSE}
gtex <- readr::read_tsv(file.path(TSR_DATA, "gtex_median_tpm.tsv"),
                        show_col_types = FALSE)

# enriched <- enrich_annotations(sites, expression_matrix = gtex)
```

Genes absent from the matrix receive `NA`, not zero, keeping "expression
unknown" distinct from "measured as zero". Pass `fill_missing_tpm = 0` to
change that. Any tissue column name may then be given as
`primary_tissue`, so the names in your matrix determine the choices
available.

A different atlas works equally well — the function requires only the
`gene_symbol`-plus-tissues shape. A programme with in-house RNA-seq from the
tissues its formulation actually reaches is better served by that than by a
population atlas.

## Step 5: cross-species orthologs

`species_match()` needs a table keyed by `human_symbol`, with two columns per
species: `{species}_ortholog_type` and `{species}_identity_pct`. Without one,
the bundled `ortholog_table_example` is used — again 20 genes of simulated
values, and genes outside it score zero, which drags any real result towards
zero.

The table is a BioMart download. Only the genes you actually score are
needed, which keeps it small: the published analyses covered 2,143 genes in a
201 KB file. Query in batches, since a single request carrying thousands of
symbols exceeds the server's URL limit.

```{r ortholog-download, purl = FALSE}
library(biomaRt)

species <- c(mouse = "mmusculus", rat = "rnorvegicus",
             cyno  = "mfascicularis", rhesus = "mmulatta",
             rabbit = "ocuniculus", dog = "clfamiliaris")

# The genes to score: the critical off-targets from your own run. Run this
# after the scanning and enrichment steps below have produced `enriched`.
critical <- enriched |>
  dplyr::filter(.data$cancer_gene_v2 | .data$AE_gene | .data$immune_gene_immport) |>
  dplyr::distinct(human_symbol = toupper(.data$gene_name)) |>
  dplyr::arrange(.data$human_symbol)

mart <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl")

fetch_species <- function(prefix, genes, batch = 250L) {
  idx <- split(seq_along(genes), ceiling(seq_along(genes) / batch))
  out <- lapply(idx, function(i) getBM(
    attributes = c("external_gene_name",
                   paste0(prefix, "_homolog_orthology_type"),
                   paste0(prefix, "_homolog_perc_id")),
    filters = "hgnc_symbol", values = genes[i], mart = mart))
  z <- dplyr::bind_rows(out)
  names(z) <- c("human_symbol", "ortholog_type", "identity_pct")
  # BioMart may return co-orthologs: prefer one2one, else highest identity.
  z |>
    dplyr::mutate(human_symbol = toupper(.data$human_symbol),
                  ortholog_type = sub("^ortholog_", "", .data$ortholog_type)) |>
    dplyr::arrange(.data$human_symbol,
                   dplyr::desc(.data$ortholog_type == "one2one"),
                   dplyr::desc(.data$identity_pct)) |>
    dplyr::slice_head(n = 1L, by = "human_symbol")
}

ortholog <- critical
for (sp in names(species)) {
  z <- fetch_species(species[[sp]], critical$human_symbol)
  names(z)[2:3] <- paste0(sp, c("_ortholog_type", "_identity_pct"))
  ortholog <- dplyr::left_join(ortholog, z, by = "human_symbol")
}
# Genes BioMart could not resolve: mark explicitly, they score zero.
for (sp in names(species)) {
  col <- paste0(sp, "_ortholog_type")
  ortholog[[col]][is.na(ortholog[[col]])] <- "none"
}

readr::write_csv(ortholog, file.path(TSR_DATA, "ortholog_biomart.csv"))
```

```{r ortholog-use, purl = FALSE}
ortholog <- readr::read_csv(file.path(TSR_DATA, "ortholog_biomart.csv"),
                            show_col_types = FALSE)
species <- species_match(enriched, ortholog_table = ortholog,
                         candidates = c("mouse","rat","cyno","rhesus","rabbit","dog"))
```

Two points on interpretation. Homology attributes are not uniform across
Ensembl releases: release 113 does not expose human-to-rat or
human-to-rabbit homology, so the published analyses scanned against 113 while
downloading homology from Ensembl Genes 116. Record both versions if they
differ. And conservation is assessed per gene, at the level of ortholog
identity — not at the individual binding site, so a conserved gene whose
seed-matching window differs between species still scores highly.

For a quick look at a single siRNA against the 3' UTR (which dominates
the off-target landscape in practice), passing only the 3' UTR `annodb`
is sufficient. To reproduce the four-class risk distribution shown in
the reference cohort plots, all three feature classes need to be scanned
and their results concatenated (Section 5.2 shows the pattern).

# The reference cohort

The package ships with `reference_set`, a compact summary of the
off-target landscape of 94 published human siRNAs curated from the
MIT/ICBP siRNA Database. All guides are 19 nt, normalized with
`normalize_guide()`, which strips the written-out `dTdT` overhang and caps
the length from the 3' end. It contains per-siRNA aggregate metrics,
per-dimension percentile thresholds for normalization, a
species × siRNA conservation matrix, and a per-siRNA risk-score
distribution. It does not contain the underlying per-site hit table —
that material is regenerated from `data-raw/build_reference_set.R`
when needed.

```{r reference, purl = FALSE}
str(reference_set, max.level = 1)

reference_set$n_siRNAs            # 94
head(reference_set$metadata, 3)   # siRNA_name, guide, target_gene, ...
head(reference_set$metrics, 3)    # count endpoints + composite_score
reference_set$tier_thresholds     # min/max/Q25/Q50/Q75 per dimension
```

## Preparing your own siRNA

`annotate_sites()` requires the **guide (antisense)** strand, because it
searches the transcriptome for `reverseComplement(sequence)`. Passing a
sense strand does not raise an error — it returns a plausible-looking
table of seed-only hits with no on-target match. Run your sequences
through `normalize_guide()` first, both to fix the strand and to land on
the same 19 nt convention as the reference cohort:

```{r normalize, eval = FALSE, purl = FALSE}
# A published 21-mer guide: capped to 19 nt from the 3' end
normalize_guide("UCCAUAACUUCUUGCUAAGUC")
#> "UCCAUAACUUCUUGCUAAG"

# Only the sense strand to hand? Ask for the reverse complement
normalize_guide("GUUUUCACUCCAGCUAACAdTdT", input_strand = "sense")
#> "UGUUAGCUGGAGUGAAAAC"
```

Sanity check after scanning: a correctly oriented guide produces a
`full_complementarity` hit with `n_mismatch == 0` on its own target gene.
If no such row appears, the strand is probably reversed.

The reference set is the basis for all functions that compare a new
siRNA against a meaningful background distribution: `rank_sirna()` uses
its `tier_thresholds` for normalization, and several plotting helpers
overlay reference data behind the user's results (Section 6).

# Three input modalities

TargetSureR accepts three distinct input types and dispatches each to a
different backend.

## Mode A, single siRNA: from a guide sequence

When the investigator has a candidate guide sequence and wants to know
its genome-wide off-target profile, the input is a list with two named
elements. The simplest case scans only the 3' UTR:

```{r mode-a-single, purl = FALSE}
sites_3utr <- annotate_sites(
  input    = list(siRNA_name = "Patisiran",
                  sequence   = "AUGGAAUACUCUUGGUUAC"),
  mode     = "sequence",
  annodb   = annodb_3utr,
  save_raw = "patisiran_out/raw"
)
sites_5utr <- annotate_sites(
  input    = list(siRNA_name = "Patisiran",
                  sequence   = "AUGGAAUACUCUUGGUUAC"),
  mode     = "sequence",
  annodb   = annodb_5utr,
  save_raw = "patisiran_out/raw"
)
sites_cds <- annotate_sites(
  input    = list(siRNA_name = "Patisiran",
                  sequence   = "AUGGAAUACUCUUGGUUAC"),
  mode     = "sequence",
  annodb   = annodb_cds,
  save_raw = "patisiran_out/raw"
)
sites_combined=rbind(sites_3utr, sites_5utr, sites_cds)
dim(sites_combined)
```

Internally, the function runs two scans on the transcriptome contained
in `annodb`: a seed-based scan via `SeedMatchR` (using the canonical
mer7m8 seed, positions 2–8 of the guide), and a full-length scan via
`Biostrings::vmatchPattern` allowing up to three mismatches. Hits from
both scans are merged with a deduplication priority that favors
full-complementarity matches over seed-only matches.

The returned tibble has thirteen columns: ten core fields shared with
Mode B (`siRNA_name`, `siRNA`, `strand`, `transcript`, `gene_name`,
`biotype`, `position`, `region`, `match_type`, `risk_score`) plus three
sequence-specific fields:

| Column | Description |
|---|---|
| `target_site` | mRNA subsequence at the hit, RNA notation, same length as the guide |
| `n_mismatch` | mismatches between guide reverse-complement and `target_site` |
| `alignment` | three-line text alignment (guide / target / match line) |

A separate file `raw_<siRNA_name>.tsv` is written under `save_raw/` and
preserves every individual hit from each scan engine, before merging.
This is the audit trail for the analysis.

## Mode A, multiple siRNAs: candidate-set screening

To screen N candidate siRNAs simultaneously, iterate over a list of
inputs and concatenate the results before any downstream step. The
example below scans each candidate against all three feature classes
(3' UTR, 5' UTR, CDS) so that the resulting hit table covers the full
range of risk-score values:

```{r mode-a-batch, purl = FALSE}
sirna_list <- list(
  list(name = "cand_01", seq = "AUGGAAUACUCUUGGUUAC"),
  list(name = "cand_02", seq = "UCCUGUUGCUGAGUAUCAU"),
  list(name = "cand_03", seq = "GUACCUGGAUAUACUAAGA")
)

feat_dbs <- list(`3UTR` = annodb_3utr,
                 `5UTR` = annodb_5utr,
                 `cds`  = annodb_cds)

sites_list <- list()
for (s in sirna_list) {
  per_feat <- lapply(names(feat_dbs), function(feat) {
    annotate_sites(
      input        = list(siRNA_name = s$name, sequence = s$seq),
      mode         = "sequence",
      annodb       = feat_dbs[[feat]],
      feature_type = feat)
  })
  sites_list[[s$name]] <- dplyr::bind_rows(per_feat)
}
sites <- dplyr::bind_rows(sites_list)
```

A single concatenated `sites` object then feeds the rest of the
pipeline. Combining the per-siRNA results before enrichment ensures
that downstream summaries — particularly `rank_sirna()` —
operate over the entire batch rather than per-call. For a single-siRNA
analysis where a 3' UTR scan is sufficient, the inner loop over
`feat_dbs` collapses to a single `annotate_sites()` call.

## Mode B: from pre-computed (transcript, position) coordinates

When upstream pipelines have already produced a list of off-target
coordinates, Mode B annotates them using a local `EnsDb` snapshot, with
optional fallback to the Ensembl REST API for transcript identifiers
released after the local snapshot. The input is a data frame:

```{r mode-b, purl = FALSE}
input_b <- data.frame(
  siRNA_name = c("hsiR22", "hsiR22", "hsiR21"),
  transcript = c("ENST00000426385.4",
                  "ENST00000335508.11",
                  "ENST00000380152.7"),
  position   = c(379L, 3136L, 500L)
)

sites_2 <- annotate_sites(
  input              = input_b,
  mode               = "position",
  ensdb_path         = ensdb_path,   # from step 2
  rest_fallback      = TRUE,
  with_lookup_source = TRUE
)
```

Because the input contains no nucleotide information, Mode B does not
populate `target_site`, `n_mismatch`, or `alignment`. Setting
`with_lookup_source = TRUE` adds an extra column that records whether
each transcript was resolved from the local EnsDb (`"ensdb"`), the REST
fallback (`"rest"`), or could not be resolved at all (`"unresolved"`).

# Downstream analysis

Once `sites` exists in either form, the rest of the pipeline is
identical across input modes.

## Annotation enrichment

```{r enrich, purl = FALSE}
# With your own expression matrix from step 4 (what an analysis should do)
enriched <- enrich_annotations(sites, expression_matrix = gtex)

# Omitting it falls back to the 20-gene simulated example: fine for a
# first look at the table's shape, not for interpreting expression.
enriched_demo <- enrich_annotations(sites)
attr(enriched_demo, "genes_missing_expression")   # most of them
```

This step joins three curated gene-list flags (`AE_gene`,
`cancer_gene_v2`, `immune_gene_immport`) and one column per tissue to each
off-target site, producing the enriched table that is the basis for all
scoring. The gene lists are bundled in full, so those flags are correct
either way; only the expression columns depend on step 4. The
`genes_missing_expression` attribute reports how many off-target genes the
matrix did not cover — with the example matrix this is nearly all of them,
which is the signal that the expression columns cannot be interpreted.

## Composite risk ranking

```{r rank, purl = FALSE}
ranking <- rank_sirna(enriched)            # default: reference = "builtin"
ranking <- rank_sirna(enriched, reference = "self")  # within-batch only
```

`rank_sirna()` summarizes seven per-siRNA metrics (high-risk hits in
critical genes, total critical-gene hits, full-complementarity in CDS,
total full-complementarity, total CDS hits, total high-risk hits, and
seed-region hits in 3' UTRs), normalizes them against either the
built-in reference cohort or the input batch itself, applies the
weights returned by `default_weights()`, and bins the resulting
composite score into Low / Medium / High / Critical tiers.

A single new siRNA processed under `reference = "builtin"` therefore
receives an interpretable absolute score — its position relative to the
94-siRNA reference distribution. The same input under
`reference = "self"` would yield a composite of zero, because there is
no within-batch variance.

## Tissue safety profile

```{r tissue, purl = FALSE}
tissue <- tissue_safety(enriched, primary_tissue = "Nerve_Tibial")
head(tissue, 10)
```

This produces a per-gene `bio_attention` score that combines four
components: normalized tissue TPM, normalized maximum risk-score,
normalized rate of full-complementarity in CDS, and a binary critical-
gene indicator.

## Animal-model recommendation

```{r species, purl = FALSE}
species <- species_match(
  enriched,
  ortholog_table = ortholog,          # from step 5
  candidates = c("mouse","rat","cyno","rhesus","rabbit","dog"))
species$species_score
species$recommendation
```

For each siRNA, the ortholog conservation score is computed across
critical-gene hits in each candidate species, and the species with the
highest score is reported. Each unique gene contributes once, so genes
carrying several sites are not weighted more heavily.

Omitting `ortholog_table` falls back to the 20-gene simulated example, in
which genes outside those 20 resolve to no ortholog and score zero. The
resulting scores reflect the example table's coverage rather than real
conservation, so step 5 is a prerequisite for using this output. Species
recommendation is in any case one input among several: exposure, historical
database and lesion relevance also bear on the decision.

## One-call report bundle

```{r report, purl = FALSE}
generate_report(
  enriched,
  output_dir     = "patisiran_out",
  target_sirnas  = "Patisiran",
  primary_tissue = "Nerve_Tibial",
  species_result = species,
  with_figures   = TRUE
)
```

The report bundle writes five CSV tables and ten figure files (six
standard plots and four reference-cohort comparisons) to a target
directory.

# Plotting helpers

Each helper returns a `ggplot` object that can be saved with
`ggplot2::ggsave()`, modified with additional layers, or composed with
other plots through e.g. `cowplot::plot_grid()`.

## Standard plots

```{r plots-standard, purl = FALSE}
plot_ranking_bar(ranking)
plot_risk_distribution(enriched)
plot_gene_attention(tissue, top_n = 30)
plot_species_heatmap(species)
plot_tissue_heatmap(enriched, tissue)
plot_critical_bubble(enriched, tissue)
```

## Reference-cohort comparison plots

These four helpers overlay or combine the user's results with the
built-in 94-siRNA reference cohort.

```{r plots-reference, purl = FALSE}
# Composite-score violin: 94 reference siRNAs as gray jitter,
# user siRNA as a colored diamond, Q25/Q50/Q75 annotated as dashed lines
plot_ranking_bar(ranking, reference = "builtin")

# All-cohort risk-score stacked bars: 94 reference + user side-by-side,
# user bar outlined and labeled in red
plot_risk_distribution(enriched, reference = "builtin")

# Per-species conservation distribution (violin + jitter + diamond)
plot_species_vs_reference(species, reference = "builtin")

# Per-species percentile rank of the user siRNA against reference
plot_species_percentile(species, reference = "builtin")
```

`generate_figures()` produces all standard and reference-comparison
figures in one call:

```{r generate-figures, purl = FALSE}
generate_figures(
  enriched, ranking, tissue, species,
  output_dir     = "patisiran_out/figures",
  formats        = c("pdf","png"),
  with_reference = TRUE
)
```

# An optional annotation step for sequence-only inputs

When an investigator supplies only a guide sequence and a project name,
the resulting `sites` table inherits the `gene_name` column directly
from `annodb$gtf`, which records the longest annotated transcript per
gene for the loaded feature class. In some downstream contexts —
particularly when combining results across feature classes (3' UTR,
5' UTR, CDS) — it is useful to verify these gene assignments against a
separate biotype query, or to attach RefSeq accessions and HGNC symbols.
A minimal annotation step looks as follows:

```{r reannotate, purl = FALSE}
library(ensembldb)
library(AnnotationFilter)

unique_tx <- unique(sites$transcript)
meta <- transcripts(
  edb,
  filter      = TxIdFilter(unique_tx),
  columns     = c("tx_id", "gene_name", "tx_biotype",
                  "gene_biotype", "seq_name"),
  return.type = "data.frame"
)
sites$gene_biotype <- meta$gene_biotype[match(sites$transcript, meta$tx_id)]
sites$chr          <- meta$seq_name[match(sites$transcript, meta$tx_id)]
```

When transcript identifiers from a newer Ensembl release are not
present in the local EnsDb, the same lookup can be obtained at runtime
from the Ensembl REST API by setting `rest_fallback = TRUE` in
`annotate_sites()`. The augmented columns can then be passed forward to
`enrich_annotations()` without further changes.

# Using a different reference cohort

The bundled `reference_set` derives from the MIT/ICBP siRNA Database, whose
144 human-reactive entries were filtered to the 94 synthetic duplexes,
excluding 50 long-hairpin shRNA constructs. Two properties bound what its
percentiles mean: the deposits are cell-biology reagents weighted towards
cancer signalling, and the database was last updated in January 2009. A
percentile therefore says a candidate is typical or atypical *among those
reagents* — which is a different statement from safe. A programme working on
conjugated hepatic or ocular siRNAs is better served by its own cohort.

Substituting one means running the same pipeline over your own guides and
passing the result as a custom reference:

```{r custom-reference, purl = FALSE}
# 1. Scan your cohort's guides exactly as in "Mode A, multiple siRNAs" above
#    and enrich them, giving `cohort_enriched`. Normalise within the cohort:
cohort_ranking <- rank_sirna(cohort_enriched, reference = "self")

# 2. Derive the per-dimension min/max/quartiles that define the new scale.
dims <- names(default_weights())
tier_thresholds <- do.call(rbind, lapply(dims, function(d) {
  v <- cohort_ranking[[d]]
  data.frame(dimension = d, min = min(v), max = max(v),
             q25 = quantile(v, 0.25, names = FALSE),
             q50 = quantile(v, 0.50, names = FALSE),
             q75 = quantile(v, 0.75, names = FALSE))
}))

my_reference <- list(metrics = cohort_ranking,
                     tier_thresholds = tier_thresholds,
                     n_siRNAs = nrow(cohort_ranking))

# 3. Place your candidates against that cohort instead of the bundled one.
ranking <- rank_sirna(enriched, reference = "custom",
                      reference_set = my_reference)
```

Note `reference = "self"` at step 1: the default `"builtin"` would normalise
your cohort against the bundled one and put the scores on a foreign scale.
`reference_set` needs at minimum a `tier_thresholds` data frame with those six
columns; supplying `metrics` as well lets the reference-overlay plots show
your cohort's distribution rather than the bundled one.

The script that produced the bundled set, `data-raw/build_reference_set.R`,
is in the source repository but excluded from the installed package, and it
reads a curated sequence table that is not redistributed with it. Treat it as
a record of how the bundled cohort was made rather than a runnable recipe.
End-to-end runtime was about an hour on one workstation, dominated by
`SeedMatchR::SeedMatchR()` on the larger feature classes.

# Session information

```{r session, eval = TRUE}
sessionInfo()
```
